The response theory for the system under a non-equilibrium driving

### 2013-06-19 writedown/linear.response/paper.tex
add numerical result for optimal forcing. better notation for the magnitude of F_e

This time, we consider the following Langevin equation:
\begin{align}\label{eqn:langevin-1b}
  \dot{\vect q} & = \nabla_{\vect p}\mh(\vect q,\vect p), \\\label{eqn:langevin-2b}
  \dot{\vect p} & =- \nabla_{\vect q}\mh(\vect q,\vect p)
  + (k_0 F_e(t)+\eps \Delta F_e(t)) \vect D(\vect q)
  - \gamma\vect p
  + \sigma\dot{\vect W},
\end{align}
such that for $\eps=0$ we get a driven system that is further
perturbed by the addition time-dependent driving force $\eps \Delta
F_e(t) \vect D(\vect q)$. Let $\vect X^\eps_t$ denote the associated solution process, and
$\mathcal P_\eps$ the probability measure of the trajectories
generated by driven Langevin
dynamics~\eqref{eqn:langevin-1b}--\eqref{eqn:langevin-2b}.

@@

\section{Optimal Forcing}

@@

Now we consider the optimization problem
\[
O_{max}(T) = \max_{F_e\in \mathcal{F}} O[F_e](T),
\]
where $O[F](T)$ is the expectation value of the path observable $\mathcal{O}$ that results from \eqref{eqn:langevin-1b}--\eqref{eqn:langevin-2b} with $F_e=F$ and $\eps=0$.
With the gradient being computed one can solve the optimization problem by means of gradient descent in coefficient space,
\[
a^{(i+1)}=a^{(i)} -\gamma \frac{\partial O(T)}{\partial a}.
\]

@@

\begin{figure}
  \centering
  \includegraphics[width=0.95\textwidth]
  {figs/fig-tilt-str2-2d-more.eps}
  \caption{The plot of $\Delta\mt_\tau f(\vect x,t)$ under
    perturbation with strength $\eps = 2$.  From left to right the
    columns present results at time $t = 0$, 10, 15 and
    20~\textsf{ps}.  The 1st row: the direct non-equilibrium
    simulation. The 2nd row: the results of traditional equilibrium
    linear response theory.  The 3rd row: the results of the response
    theory developed by this work.  We calculate response terms up to
    the second order from an equilibrium simulation (i.e. letting $k_0
    = 0$ and $\eps = 0$ in Eq.~\eqref{eqn:langevin-2b}).
  }
  \label{fig:tmp2}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.95\textwidth]
  {figs/fig-tilt-str4-2d-more.eps}
  \caption{The plot of $\Delta\mt_\tau f(\vect x,t)$  under perturbation of
    $\eps = 4$. From left to right the
    columns present results at time $t = 0$, 10, 15 and
    20~\textsf{ps}.
    The first three rows use the same algorithm as Fig.~\ref{fig:tmp2}.    
    The 4th row: the results of the response theory developed by this work.
    Instead of calculating the responses from the equilibrium simulation,
    here we calculate the linear response from a non-equilibrium simulation
    of strength 2 (i.e. $k_0 = 2$ and $\eps = 4$
    in Eq.~\eqref{eqn:langevin-2b}).
    Therefore, the effective perturbation in this case is only
    $\eps = 2$.
  }
  \label{fig:tmp3}
\end{figure}

We test the idea firstly by a one-dimensional model system: one particle in a
tilting double-well potential. For convenience, we let the mass of the
particle to be 1~\textsf{amu},
and the friction coefficient to be $1\,\textsf{ps}^{-1}$.
The unperturbed
Hamiltonian of the system is given by:
\begin{align}
  \mh (\vect p, \vect q) = \frac 12 \vect p^2 + U(\vect q) 
\end{align}
with potential
\begin{align}
  U(\vect q) = \frac12 k (\vect q^2 - a^2)^2
\end{align}
Here $k = 8$~$\textsf{kJ} / (\textsf{mol nm}^4)$, and $ a = 1\ \textsf{nm}$.
At room temperature $300\ \textsf{K}$, $k_BT = 2.48$~\textsf{kJ/mol}.
The red line in Fig.~\ref{fig:tmp1} presents the unperturbed potential.
%$\textsf{kJ} / (\textsf{mol nm}^4)}$
The perturbation is given by
\begin{align}
  \vect D(\vect q) = -\nabla_{\vect q} V(\vect q) = 1
\end{align}
Here $V(\vect q) = -\vect q$   effectively tilts the original
potential $U(\vect q)$ (see the gree and blue lines in Fig.~\ref{fig:tmp1}).
The strength of $F_e(t)$ is
set to be 1, when the time $t$ is larger than some warm-up time $t_c$.
When $t \leq t_c$, the strength of $F_e(t)$ is linearly growing,
i.e. $ F_e(t) = t/t_c$.
We use the value $t_c = 20$~\textsf{ps}, so that the perturbation increases slow
enough: the typical decaying time scale of the correlation function
in~\eqref{eqn:eq-lr} is roughly 3~\textsf{ps}.
We choose $\tau = 1$~\textsf{ps}, and
% We choose $\mt_\alpha$
% to be the propagator of the Langevin dynamics at temperature $150$
% \textsf{K} with $\alpha = 1$ \textsf{ps}.
consider the stability of the
distribution $f(\vect x, t)$ at time $t$: $\Delta\mt_\tau f(\vect x, t)$.
The larger
this value, the more stable the core sets are.

@@

In Fig.~\ref{fig:tmp3}, as the
perturbation grows stronger, the left well becomes less stable, while
the right well becomes more stable.  When the perturbation is as strong as
$\eps = 4$, the linear response result (2nd row) is no
longer precise:
At $t\geq 20$ \textsf{ps}, the left well basically
disappears, however, the linear response calculation presents
artificial structure (that is a small artificial
stable region accompanied by a small unstable region)
at the position of the left well.
The second order response (3rd row) improves the accuracy a little, 
however, it also includes artificial finer-grained oscillation in the profile.
At the same time, the numerical uncertainty is higher, so the figure
looks noisy.
In the 4th row of Fig.~\ref{fig:tmp3}, we present the linear response result
that calculate from 
a intermediate
non-equilibrium simulation with driving force of $k_0 = 2$.
Notice that the reference simulation in this case is also a
non-equilibrium simulation, so the effective strength of perturbation
is reduced to $\eps = 2$. According to Fig.~\ref{fig:tmp1}, this can be
considered as a small perturbation.
A substantial improvement is achieved: the profile
is correctly calculated, and the statistical error is small.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.95\textwidth]
  {figs/fig-split-str2-2d-more.eps}
  \caption{
    The plot of $\Delta\mt_\tau f(\vect x,t)$  under perturbation of
    $\eps = 2$. From left to right the
    columns present results at time $t = 0$, 5, 10 and
    20~\textsf{ps}.
    The 1st row: the direct non-equilibrium
    simulation. The 2nd row: the results of traditional equilibrium
    linear response theory.  The 3rd row: the results of the response
    theory developed by this work.  We calculate response terms up to
    the second order from an equilibrium simulation (i.e. letting $k_0
    = 0$ and $\eps = 2$ in Eq.~\eqref{eqn:langevin-2b}).
    The 4th row: the results of the response theory developed by this work.
    Instead of calculating the responses from the equilibrium simulation,
    here we calculate the linear response from a non-equilibrium simulation
    of strength 2 (i.e. $k_0 = 1$ and $\eps = 1$
    in Eq.~\eqref{eqn:langevin-2b}).
    Only the linear term of response is calculated.
    Notice, the effective perturbation in this case is only
    $\eps = 1$.
  }
  \label{fig:tmp6}
\end{figure}

@@

In this example, we also consider two different strengths of
perturbations. Please see the green and blue lines in Fig~\ref{fig:tmp4} for
perturbations of $\eps = 1$ and $\eps = 2$, respectively.
Numerical results show that the perturbation $\eps=1$ can be considered to be
relatively small. Since the phenomenon is basically the same as
Fig.~\ref{fig:tmp2}, we do not present them here.
Fig.~\ref{fig:tmp6} presents the numerical results of perturbation $\eps =
2$.  Only with a linear response calculated from the equilibrium
simulation (2nd row), the stability of the two new core sets are too
strong, and the unstable region between them is over estimated, comparing with
the brute force simulation presented on the first row.
We present the  second order response results on the 3rd row, which is
also calculated from the equilibrium simulation. In this case, the statistical
error is too large, so that
the profile of $\mt_\tau f$ is indeed too noisy to draw any useful information.
On the 4th row of Fig.~\ref{fig:tmp6} presents the linear response starting
from a non-equilibrium simulation with external driving $k_0 = 1$.
In this case the effective strength of the perturbation to the reference
simulation is $\eps = 1$, which is considered to be relatively small.
The numerical results is satisfactorily consistent with the brute force
non-equilibrium simulation.

>>>>

This time, we consider the following Langevin equation:
\begin{align}\label{eqn:langevin-1b}
  \dot{\vect q} & = \nabla_{\vect p}\mh(\vect q,\vect p), \\\label{eqn:langevin-2b}
  \dot{\vect p} & =- \nabla_{\vect q}\mh(\vect q,\vect p)
  + (F_e(t)+\eps \Delta F_e(t)) \vect D(\vect q)
  - \gamma\vect p
  + \sigma\dot{\vect W},
\end{align}
such that for $\eps=0$ we get a driven system that is further
perturbed by the addition time-dependent driving force $ \Delta
F_e(t) \vect D(\vect q)$. Let $\vect X^\eps_t$ denote the associated solution process, and
$\mathcal P_\eps$ the probability measure of the trajectories
generated by driven Langevin
dynamics~\eqref{eqn:langevin-1b}--\eqref{eqn:langevin-2b}.

@@

\subsection{Optimal Forcing}

@@

Now we consider the optimization problem
\[
O_{min}(T) = \min_{F_e\in \mathcal{F}} O[F_e](T),
\]
where $O[F](T)$ is the expectation value of the path observable $\mathcal{O}$ that results from \eqref{eqn:langevin-1b}--\eqref{eqn:langevin-2b} with $F_e=F$ and $\eps=0$.
With the gradient being computed one can solve the optimization problem by means of gradient descent in coefficient space,
\[
a^{(i+1)}=a^{(i)} -\gamma \frac{\partial O(T)}{\partial a}.
\]

@@

\begin{figure}
  \centering
  \includegraphics[width=0.95\textwidth]
  {figs/fig-tilt-str2-2d-more.eps}
  \caption{The plot of $\Delta\mt_\tau f(\vect x,t)$ under
    perturbation with strength $\eps = 2$.  From left to right the
    columns present results at time $t = 0$, 10, 15 and
    20~\textsf{ps}.  The 1st row: the direct non-equilibrium
    simulation. The 2nd row: the results of traditional equilibrium
    linear response theory.  The 3rd row: the results of the response
    theory developed by this work.  We calculate response terms up to
    the second order from an equilibrium simulation (i.e. letting $k_e
    = 0$ and $\eps = 0$ in Eq.~\eqref{eqn:langevin-2b}).
  }
  \label{fig:tmp2}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.95\textwidth]
  {figs/fig-tilt-str4-2d-more.eps}
  \caption{The plot of $\Delta\mt_\tau f(\vect x,t)$  under perturbation of
    $\eps = 4$. From left to right the
    columns present results at time $t = 0$, 10, 15 and
    20~\textsf{ps}.
    The first three rows use the same algorithm as Fig.~\ref{fig:tmp2}.    
    The 4th row: the results of the response theory developed by this work.
    Instead of calculating the responses from the equilibrium simulation,
    here we calculate the linear response from a non-equilibrium simulation
    of strength 2 (i.e. $k_e = 2$ and $\eps = 4$
    in Eq.~\eqref{eqn:langevin-2b}).
    Therefore, the effective perturbation in this case is only
    $\eps = 2$.
  }
  \label{fig:tmp3}
\end{figure}

We test the idea firstly by a one-dimensional model system: one particle in a
tilting double-well potential. For convenience, we let the mass of the
particle to be 1~\textsf{amu},
and the friction coefficient to be $1\,\textsf{ps}^{-1}$.
The unperturbed
Hamiltonian of the system is given by:
\begin{align}
  \mh (\vect p, \vect q) = \frac 12 \vect p^2 + U(\vect q) 
\end{align}
with potential
\begin{align}
  U(\vect q) = \frac12 k (\vect q^2 - a^2)^2
\end{align}
Here $k = 8$~$\textsf{kJ} / (\textsf{mol nm}^4)$, and $ a = 1\ \textsf{nm}$.
At room temperature $300\ \textsf{K}$, $k_BT = 2.48$~\textsf{kJ/mol}.
The red line in Fig.~\ref{fig:tmp1} presents the unperturbed potential.
%$\textsf{kJ} / (\textsf{mol nm}^4)}$
The perturbation is given by
\begin{align}
  \vect D(\vect q) = -\nabla_{\vect q} V(\vect q) = 1
\end{align}
Here $V(\vect q) = -\vect q$   effectively tilts the original
potential $U(\vect q)$ (see the gree and blue lines in Fig.~\ref{fig:tmp1}).
The strength of $F_e(t)$ is
set to be $k_e$, when the time $t$ is larger than some warm-up time $t_c$.
When $t \leq t_c$, the strength of $F_e(t)$ is linearly growing,
i.e. $ F_e(t) = k_e\cdot t/t_c$.
We use the value $t_c = 20$~\textsf{ps}, so that the perturbation increases slow
enough: the typical decaying time scale of the correlation function
in~\eqref{eqn:eq-lr} is roughly 3~\textsf{ps}.
We choose $\tau = 1$~\textsf{ps}, and
% We choose $\mt_\alpha$
% to be the propagator of the Langevin dynamics at temperature $150$
% \textsf{K} with $\alpha = 1$ \textsf{ps}.
consider the stability of the
distribution $f(\vect x, t)$ at time $t$: $\Delta\mt_\tau f(\vect x, t)$.
The larger
this value, the more stable the core sets are.

@@

In Fig.~\ref{fig:tmp3}, as the
perturbation grows stronger, the left well becomes less stable, while
the right well becomes more stable.  When the perturbation is as strong as
$\eps = 4$, the linear response result (2nd row) is no
longer precise:
At $t\geq 20$ \textsf{ps}, the left well basically
disappears, however, the linear response calculation presents
artificial structure (that is a small artificial
stable region accompanied by a small unstable region)
at the position of the left well.
The second order response (3rd row) improves the accuracy a little, 
however, it also includes artificial finer-grained oscillation in the profile.
At the same time, the numerical uncertainty is higher, so the figure
looks noisy.
In the 4th row of Fig.~\ref{fig:tmp3}, we present the linear response result
that calculate from 
a intermediate
non-equilibrium simulation with driving force of $k_e = 2$.
Notice that the reference simulation in this case is also a
non-equilibrium simulation, so the effective strength of perturbation
is reduced to $\eps = 2$. According to Fig.~\ref{fig:tmp1}, this can be
considered as a small perturbation.
A substantial improvement is achieved: the profile
is correctly calculated, and the statistical error is small.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.95\textwidth]
  {figs/fig-split-str2-2d-more.eps}
  \caption{
    The plot of $\Delta\mt_\tau f(\vect x,t)$  under perturbation of
    $\eps = 2$. From left to right the
    columns present results at time $t = 0$, 5, 10 and
    20~\textsf{ps}.
    The 1st row: the direct non-equilibrium
    simulation. The 2nd row: the results of traditional equilibrium
    linear response theory.  The 3rd row: the results of the response
    theory developed by this work.  We calculate response terms up to
    the second order from an equilibrium simulation (i.e. letting $k_e
    = 0$ and $\eps = 2$ in Eq.~\eqref{eqn:langevin-2b}).
    The 4th row: the results of the response theory developed by this work.
    Instead of calculating the responses from the equilibrium simulation,
    here we calculate the linear response from a non-equilibrium simulation
    of strength 2 (i.e. $k_e = 1$ and $\eps = 1$
    in Eq.~\eqref{eqn:langevin-2b}).
    Only the linear term of response is calculated.
    Notice, the effective perturbation in this case is only
    $\eps = 1$.
  }
  \label{fig:tmp6}
\end{figure}

@@

In this example, we also consider two different strengths of
perturbations. Please see the green and blue lines in Fig~\ref{fig:tmp4} for
perturbations of $\eps = 1$ and $\eps = 2$, respectively.
Numerical results show that the perturbation $\eps=1$ can be considered to be
relatively small. Since the phenomenon is basically the same as
Fig.~\ref{fig:tmp2}, we do not present them here.
Fig.~\ref{fig:tmp6} presents the numerical results of perturbation $\eps =
2$.  Only with a linear response calculated from the equilibrium
simulation (2nd row), the stability of the two new core sets are too
strong, and the unstable region between them is over estimated, comparing with
the brute force simulation presented on the first row.
We present the  second order response results on the 3rd row, which is
also calculated from the equilibrium simulation. In this case, the statistical
error is too large, so that
the profile of $\mt_\tau f$ is indeed too noisy to draw any useful information.
On the 4th row of Fig.~\ref{fig:tmp6} presents the linear response starting
from a non-equilibrium simulation with external driving $k_e = 1$.
In this case the effective strength of the perturbation to the reference
simulation is $\eps = 1$, which is considered to be relatively small.
The numerical results is satisfactorily consistent with the brute force
non-equilibrium simulation.

\subsection{Optimal forcing}
\begin{figure}
  \centering
  \includegraphics[]{figs/fig-ctr-stat.eps}
  \caption{The optimal forcing.
    The red color indicates $\eta = 0.01$ and the
    green color indicates $\eta = 0.05$.
    In the lower plot, the dashed lines display the
    value of $P(\vect X_t)$ as the time eloves.
    The solid lines present $O[F_e](t)$.  In the upper
    plot, the solid lines present the magnitude of the
    control $F_e(t)$.
  }\label{fig:tmp7}
\end{figure}

We investigate the example of the double well potential, and
consider the problem of optimally forcing the population from the
left well to the right well. That is equal to consider the 
the following optimal forcing problem:
\begin{align}
  O_{\min}(T)& = \min_{F_e\in\mathcal F} O[F_e](T) 
\end{align}
with
\begin{align}
  O[F_e](T) &= P(\vect X_T) + \eta \int_0^T\vert F_e(t)\vect D(\vect q)\vert^2\, dt
\end{align}
where $P(\vect X_T)$ is equal to 1 when the end point of the
trajectory $\vect X_T$ is in the right well, and equal to 0 otherwise.
The integral is the ``cost'' of the control and $\eta$ is a constant
that indicates the relative magnitude of the cost. We consider the
ansatz function family $\mathcal F$ of piecewise linear functions on
$[0, T]$. 
Fig.~\ref{fig:tmp7} presents the numerical result. The magnitude of control
is shown in the upper plot, and the value of $O[F_e](T)$ and $P(\vect X_T)$
are given in the lower plot.
